---
title: "Categorization"
tutorial:
  id: "jedr-categorization"
output:
  learnr::tutorial:
    progressive: true
    theme: lumen
    highlight: espresso
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(learnr)
library(gradethis)
```

## How to make categories from data

Welcome Padawan. May your quest to become a JedR Master be fruitful. "Do or do not. There is no try."

Sometimes as journalists (or data scientist) we need to create new categories for our data. An example might be where we have a column of data that has too many different values to plot effectively. We might be able to re-categorize those values to fewer individual choices, maybe combining less important values into a generic term like "other".

Some real-worldish examples:

- You want to show race/ethnicity breakdowns in a chart or story about Hispanic representation. There are eight different values and some are quite small. You might create a column to use the values Asian, Black, Hispanic and White with all others changed to "Other".
- Perhaps you are doing a story on crime and want to highlight violent crimes vs property vs other crimes. You might create a new column that categorizes many individual charges into these buckets.

We're taking records and putting them into piles or groups based on one or more values.

> **JedR Mind Trick:** When we categorize values like this, we want to create a new column intead of overwriting an existing one. This way you can check your work. You can always remove or rename columns later.

### Goals of this training session

- To apply [`recode()`](https://dplyr.tidyverse.org/reference/recode.html) within a `mutate()` to create data categories. It's a one-to-one match, for the most part.
- To apply [`case_when()`](https://dplyr.tidyverse.org/reference/case_when.html) within a `mutate()` to create data categories. This allows more complex logic.

## Our scenario

Here at the _Galactic News Hub_, we have a dataset of all the sources we connect with on a regular basis: The `starwars` data set that comes with the `tidyverse` library.

```{r sw-setup}
library(tidyverse)
starwars
```

This data includes a variable called `species`, and our editor wants a report breaking down the species of our source list based on three broad categories: **human**, **droid**, and **other**. Our editor is hoping we can later cross reference these categories against gender and other variables within the source list.

To accomplish this, it would be helpful to have a "species_category" variable using these three values so we can group and aggregate with other columns.

Let's take a look at the species column using [`count()`](https://dplyr.tidyverse.org/reference/count.html).

```{r species-count}
starwars %>% count(species, sort = TRUE, name = "cnt")
```

You can see that **Human** and **Droid** values dominate the data. The next most frequent values is `NA`. Maybe we can convince our editor to make that category **Other/Unknown**.

## Using `recode()`

One of the easier methods to categorize data is to use [`recode()`](https://dplyr.tidyverse.org/reference/recode.html). The logic is straighforward: For this value, substitute that value.

To make clear our goal here for the _Galactic Hub News_: We want to create a new column called `species_category` that has three values: Human, Droid, Other/Unknown. And they need to be based on the existing `species` variable.

### How recode works

While is is possible to recode values into the same column, a good JedR will use the `recode()` function within a `mutate()` function, creating a new column and then filling it with the results of the recode.

```r
data %>% 
  mutate(
    new_column = recode(
      existing_column,
      "Old Value 1 from existing_column" = "New Value 1",
      "Old Value 2 from existing_column" = "New Value 2",
    )
  )
```

- This is standard usage for the `mutate()` function ... you create the new column and then set it to a value. In this case, we are setting the value to the result of the `recode()` function.
- Within `recode()`, we have to first name the column we are pulling from, then list the recoded values in order of old value, then new value.

**Note:** If you look at the [documentation](https://dplyr.tidyverse.org/reference/recode.html), you'll see that there is a **lifecycle question** about how the function works. Most R functions name a thing before assigning it: You have to have a planet before you can fill it with sentient beings. The `recode()` function breaks that convention by naming the old thing first, then describing what the new value is. We just have to get over it for now.

So, with our `starwars` data, a partial solution might look like this:

```{r recode-ex}
starwars %>% 
  mutate(
    species_category = recode(
      species,
      "Gungan" = "Other/Unknown",
      "Neimodian" = "Other/Unknown"
    )
  ) %>% # using select and slice to focus on the change
  select(name, species, species_category) %>% slice(31:40)
```

In the example above, we use `slice()` to skip down to part of the data where we can see the change. (The first 20 rows are all humans and droids.) We also used `select()` to focus on our columns of interest.

Note we've only changed _two_ of the 36 values we need to change, though. (You can see Toydarian and Dug have not been changed.) This method is very handy if you have just a few things to change.

In our case here, we _could_ list all 36 changes, but ... Use the Force, Luke.

### Using .default and .missing

There are two special values in recode() we can leverage here:

- `.default` recodes everything not specified.
- `.missing` recodes `NA` (or blank) values.

Using these, we can set Human to Human and Droid to Droid, but then everything else to Other/Unknown.

```{r recode-default}
starwars %>% 
  mutate(
    species_category = recode(
      species,
      "Human" = "Human",
      "Droid" = "Droid",
      .default = "Other/Unknown",
      .missing = "Other/Unknown"
    )
  ) %>% # using select and slice to focus on change
  select(name, species, species_category) %>% slice(31:40)

```

YLMV\*: The method you choose to use will depend on your needs. \*(Your Lightspeed May Vary)

### Check your results

> **JedR Mind Trick:** Even with these few rows, it's hard to _see all_ the changes made in your data. It is sometimes helpful to do a `count()` on your changes to check them. Just be sure to count the already-changed data, not the result you are saving into the new object.

Here we save the recoded data into a new object called `sw_species` and then run a `count()` on our two columns to check the results.

```{r recode-count}
sw_species <- starwars %>% 
  mutate(
    species_category = recode(
      species,
      "Human" = "Human",
      "Droid" = "Droid",
      .default = "Other/Unknown",
      .missing = "Other/Unknown"
    )
  )

sw_species %>% 
  count(species, species_category, sort = TRUE)

```

The `recode()` function works great if you have one-to-one changes that don't require logic. But when you have more complex needs, we need more Force.

## Padawan practice: recode

If you look at the `gender` column in the data, you'll see there are two values: **masculine** and **feminine**, but there are four records that don't have a `gender` value are listed as `NA`.

```{r gender-reveal}
starwars %>% count(gender)
```

For this quest, you will use `recode()` to fill in those blank values with **unknown**.

Note in the code we are creating a new dataframe and filling with mutated `starwars` data.

Within the recode section you'll need to note what variable you are working form, and then use one of the special helpers to fill the values as required.

At the end it prints the results.

```{r recode-exercise, exercise = TRUE}
gender_reveal <- starwars %>% 
  mutate(
    gender_clean = recode(
      # Fill in here
    )
  )

gender_reveal %>% select(name, gender, gender_clean)
```

<div id="recode-exercise-hint">
**Hint:** You should only have to set the base variable and the `.missing = ` values.
</div>

```{r recode-exercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
gender_reveal <- starwars %>% 
  mutate(
    gender_clean = recode(
      gender,
      .missing = "unknown"
    )
  )

gender_reveal %>% select(name, gender, gender_clean)
```

```{r recode-exercise-check}
grade_this_code()
```


### Quiz

(Just page through the results of your quest above.)

```{r recode-quiz, echo=FALSE}
question("Which of these characters has an unknown gender?",
  answer("Yoda"),
  answer("C-3PO"),
  answer("Captain Phasma", correct = TRUE),
  answer("Cordé")
)
```


## Using `case_when()`

The [`case_when()`](https://dplyr.tidyverse.org/reference/case_when.html) function is useful when you need to set new values based on some kind of logic.

Here we will use `case_when` to do the same thing we did above with `recode()` and then explain how it is working below.

```{r casewhen-show}
starwars %>% 
  mutate(
    species_category = case_when(
      species == "Human" ~ "Human",
      species == "Droid" ~ "Droid",
      TRUE ~ "Other/Unknown"
    )
  ) %>%
  select(name, species, species_category) %>% slice(31:40)
```

You can see the syntax works differently (once we are past the mutate).

- Within the `case_when()`, there are rules that split with `~`. On the left is a logical test, and on the right is a new value designation.
- Each rule is run in order. A value is set only once, so if a value is set in one rule, it will NOT be reset by subsequent rules. You must write your rules from the most specific to the most general.
- The last line `TRUE ~` handles all values not already changed. If left out, values not specified would be set `NA`.

### A more complex example

Unlike `recode()` where you are drawing from values in one specific column, the logic tests used with `case_when()` can be just about anything as long as they evaluate to `TRUE` or `FALSE`.

While this is a silly example, here we create a column called `size_compare` that notes if a character is "larger" or "smaller" than an average human, unless they are a droid, in which case they are labeled "mechanical".

```{r casewhen-complex-show}
starwars %>% 
  mutate(
    size_compare = case_when(
      species == "Droid" ~ "mechanical",
      height >= 170 & mass >= 80 ~ "larger",
      TRUE ~ "smaller"
    )
  ) %>% 
  select(name, size_compare)

```

Note we started with the rule for "Droid" because that is the most specific thing we needed to change. When then set the "larger" value, and the rest got "smaller."

## Padawan practice: case_when

Create a column called `main_planet` using the `homeworld` column. For each characters with a homeworld of "Tatooine" or "Naboo", label it as `TRUE` and otherwise label as `FALSE`, which will create a logical datatype column.

(This is a fill-in-the-blank thing ...)

```{r casewhen-quest, exercise = TRUE}
mainchars <- starwars %>% 
  mutate(
    main_planet_origin = case_when(
      str_detect(_________, "Tatooine|Naboo") ~ ____,
      TRUE ~ _____
    )
  )
mainchars %>% select(name, homeworld, main_planet_origin)
```


```{r casewhen-quest-solution, exercise.reveal_solution = FALSE, include = FALSE}
mainchars <- starwars %>% 
  mutate(
    main_planet_origin = case_when(
      str_detect(homeworld, "Tatooine|Naboo") ~ TRUE,
      TRUE ~ FALSE
    )
  )
mainchars %>% select(name, homeworld, main_planet_origin)
```

```{r casewhen-quest-check}
grade_this_code()
```

Page through your results to check your answers.

That's it for now.

-30-
