---
title: "ggplot basics"
tutorial:
  id: "jedr-ggplot-basics"
output:
  learnr::tutorial:
    progressive: true
    theme: lumen
    highlight: espresso
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(learnr)
```

## The very basics of ggplot

In your quest to become a JedR Master, you must start with the basics. So this tutorial will cover the first first steps and concepts of ggplot2 and the Grammar of Graphics.

It is just one of many ways to give into data visualization in R. There are many other ggplot tutorials, books and references available in Galactic Empire:

You might start with the data visualization chapter in [R for Data Science](https://r4ds.had.co.nz/data-visualisation.html) by Hadley Wickham.

And Google gives is [many, many](https://www.google.com/search?q=ggplot+tutorials) options, including these:

- [The Complete ggplot2 Tutorial](http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html)
- [Learn ggplot2](https://www.tutorialspoint.com/ggplot2/index.htm)
- [Data Visualization in R with ggplot2: A Beginner Tutorial](https://www.dataquest.io/blog/data-visualization-in-r-with-ggplot2-a-beginner-tutorial/)

There are other books, compendiums and references:

- [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html)
- [R Graphics Cookbook, 2nd edition](https://r-graphics.org/)
- [The R Graph Gallery](https://www.r-graph-gallery.com/)

## The Grammar of Graphics

An underlying theory about ggplot is this concept of the Grammar of Graphics … i.e., that you can use code (words) to describe how to build a chart layer-by-layer.

With a hat tip to Earthling [Matt Waite](http://www.mattwaite.com/), we can describe the components of the Grammar of Graphics as:

- **data**: which data you are pulling from for the chart.
- **aesthetics**: describes how to apply specific data to the plot. What is on x axis, what is on y axis, for starters.
- **geometries**: the shape the data is going to take on the graph. lines, columns, points.
- **scales**: any transformations we might make on the data.
- **layers**: how we might layer multiple geometries over top of each other to reveal new information.
- **facets**: how we might graph many elements of the same data set in the same space.

What to remember here is this: for every graphic we start with the data, and then we build a chart from it one "layer" at a time.

The best way to learn this system is to do it and explain along the way.

## The `starwars` data

We'll use dplyr's [`starwars` dataset](https://dplyr.tidyverse.org/reference/starwars.html) to work through these ggplot basics. It is loaded with the [tidyverse package](https://www.tidyverse.org/).

```{r starwars}
library(tidyverse)
starwars
```

This is the cast of characters from the space opera movies _Star Wars_.

Among the variables in `starwars` are:

- `height`: A characters' height in centimeters
- `mass`: A character's weight in kilograms

We'll use these two data points to explore the ggplot syntax.

### Verbose vs shortcuts

Like with many things in programming, ggplot has multiple paths to the same result. There is a very verbose way to describe a chart, and then there are shortcuts. We'll start with verbose.

At a minimum, a ggplot chart needs data, aesthetics (how to supply specific data) and geometries (how to show those).

```{r}
ggplot(data = starwars,
       mapping = aes(x = mass, y = height)) +
  geom_point()
```





