---
title: "Select"
tutorial:
  id: "rwdir-learnr/select"
  version: .9
output:
  learnr::tutorial:
    progressive: true
    theme: lumen
    highlight: espresso
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(learnr)
library(gradethis)

knitr::opts_chunk$set(echo = FALSE)
```

This is a work in progress. The goal is to write an example **learnr** lesson for Reporting with Data in R. This is just a draft concept.

To Do:

- [ ] Add more challenges between explanations
- [ ] Add new challenge with new data

## Learn the Force: Select

The **[select](https://dplyr.tidyverse.org/reference/select.html)** function lets you pick and choose columns (a.k.a. variables) in your data. It comes with a bunch of **selection helpers**, which we'll also cover.

### Use cases

- You might select specific columns from a dataset just to see them better. For instance, if you want to compare the values in two columns but there are many columns between them in output.
- You might want to remove unwanted columns from your data.
- You might want to change the order of the columns to put the most important information first.

As always, we start with `starwars` data. You can [review the data here](https://dplyr.tidyverse.org/reference/starwars.html). (BUT UPDATE THIS TO A LOCAL VERSION)

Let's take a quick look at the column names in `starwars`:

```{r select-intro}
starwars %>% glimpse()
```

### Select specific variables

The most basic use of `select()` is list the columns you want to keep by separating them with commas in the order you want them.

Here we use `select()` to "select" the **name**, **homeworld** and **birthyear** columns from our `starwars` data.

```{r select-show, echo=TRUE}
starwars %>% 
  select(name, homeworld, birth_year)
```

#### Padawan: Use select

1. Use `select()` to print the **name**, **species** and **homeworld** columns.

```{r select-select, exercise=TRUE}
starwars
```

```{r select-select-solution}
starwars %>% 
  select(name, species, homeworld)
```

```{r select-select-check}
grade_this_code()
```

### Drop a specific column

Or, if it is easier to remove a few columns, you could do that by putting `!` in front of a column name. Think `!` as "negate" or the "inverse of". Like "everything but this thing listed." This is a dumb example for starwars, but this removes the `name` column:

```{r select-remove, echo=TRUE}
starwars %>% 
  select(!name)
```

To negate more than one columns, use combine `c()` to list the columns. This also works with the negate `!` as `!c()`.

With this, we namethree columns to remove.

```{r select-drop-multiple, echo=TRUE}
starwars %>% 
  select(!c(height, mass, birth_year))
```

> SIDE NOTE: You can also use `-` to remove columns, but I think this is out of favor. Works like this: `select(-height, -mass)`.

#### Padawan: Drop columns

1. Using the starwars data, use `select()` and the selection helper `contains()` to display the **name** column and **all the columns with "color" in their name**. 

```{r quest, exercise=TRUE}
starwars
```

```{r quest-solution, include=FALSE}
starwars %>% 
  select(name, contains("color"))
```

```{r quest-check, include=FALSE}
grade_this_code()
```

### Select based on a range

You can use the ordinal number of the columns (the first starts with 1), including with a range using `:`. So this gets you the first four columns in the tibble:

```{r select-range, echo=TRUE}
starwars %>% 
  select(1:4)
```

But you can also use that range with well-formatted columns names. This also gets the first four columns, but really it is selecting everything between `name` and `hair_color`.

Here is a named range example:

```{r select-range-name, echo=TRUE}
starwars %>% 
  select(name:hair_color)
```


#### Padawan: Remove a range

1. Remove the **birth_year** column along with **films-through-starships columns**. (i.e., use a named range).

```{r select-drop-quest, exercise=TRUE}
starwars
```

```{r select-drop-quest-solution}
starwars %>% 
  select(!c(name, birth_year, films:starships))
```

```{r select-drop-quest-check}
grade_this_code()
```


### Selection helpers

There are a bunch of helpers to select columns as well. See the list in the [select docs](https://dplyr.tidyverse.org/reference/select.html) but here are some examples of more useful ones.

There are different `*_with` operators where you can pluck out columns based on part of their name. This is part of the reason a JedR Knight is anal retentive about naming columns.

We can use `ends_with()` to fetch columns that have part of a name, like this:

```{r select-ends, exercise=TRUE}
starwars %>% 
  select(ends_with("_color"))
```

There is also `starts_with()` and `contains()` that works similarly.

Sometimes you just want to move a column or two to the front of the tibble but then leave the rest. The function `everything()` is good for that as it includes every column not already called.

This example selects all the "color" columns first, then uses `everything()` to include the rest.

```{r select-everything, exercise=TRUE}
starwars %>% 
  select(ends_with("color"), everything())
```


## Use the Force: Select

