---
title: "JedR Trial 1"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse) 
library(gradethis)

starwars <- starwars
gradethis_setup()

praise_list <- c("Great job you have done!", "Nice work. You have controlled your fear.", "Wow! The force is with you!", "Great job! Mind tricks don't work on you!")

#random_sw_praise <- function(){sample(praise_list, 1)}
```

## JedR Trail 1: The Tidyverse

Congratulations! You have recently been hired as a **data jedi** for the *Galactic News Hub*, the premiere news organization for the Star Wars Universe. Your first task: apply the `tidyverse` skills that you learned this week to a dataset of Star Wars characters.

### 1. Previewing our data

For this exercise, we will using a dataset called `starwars`. This dataset comes with `tidyverse`, so if you have loaded `tidyverse` into your R environment using `library()`, the `starwars` dataset should be there! As you can see below, `starwars` contains some information about Star Wars characters, like their name and height.

```{r starwarsintro, eval = FALSE}
library(tidyverse)
starwars 
```

First, let's learn about the `starwars` dataset. The first thing we'll want to know is how many characters are in the `starwars` dataset. What function would you use to *glimpse* the data to find that answer?

```{r glimpseexercise, exercise = TRUE}

```

<div id="glimpseexercise-hint">
**Hint:** Check out the `glimpse()` function, which we talk about in [Chapter 3](https://utdata.github.io/rwdir/counts-import.html#glimpse-the-data)
</div>

```{r glimpseexercise-solution}
glimpse(starwars)
```

```{r glimpseexercise-check}
grade_this({
  if(identical(.result,.solution)){ 
    pass(sample(praise_list, 1))
  }
  fail()
})

```

(Don't forget to **Submit Answer**. You _may_ need to click the "Run Code" button again to answer the next question.)

Just as you submitted the answer, your humanoid robot editor (J-327) pops into your office and asks you how many people (characters) are in your dataset. What is your answer?

```{r quiz-glimpse, echo = FALSE}
quiz(
  question("How many Star Wars characters are in the dataset",
    answer("14"),
    answer("87", correct = TRUE),
    answer("41"),
    answer("Unknown")
  )
)
```

### 2. Focusing on variables {data-progressive=TRUE}

Now that we know how to *glimpse* into our dataset, let's work on *select*ing specific variables of interest. Let's say the *Galactic News Hub* has asked you to write a story about height differences across different species. What function would you use to *select* only the following variables in `starwars`: `name`, `height`, `homeworld`, and `species`?

```{r selectexercise, exercise = TRUE}

```

<div id="selectexercise-hint">
**Hint:** I recommend checking out the `select()` function, which we talk about in [Chapter 3](https://utdata.github.io/rwdir/counts-import.html#selecting-columns)
</div>

```{r selectexercise-solution}
starwars %>%
  select(
    name,
    height,
    homeworld,
    species
  )
```

```{r selectexercise-check}
grade_this({
  if(identical(.result, .solution)){
    pass(sample(praise_list, 1))
  }
  fail()
})
```

### 3. Arranging our Data {data-progressive=TRUE}

```{r arrangeexercise-setup, include=FALSE}
starwars <- starwars %>%
  select(
    name,
    height,
    homeworld,
    species
  )
```

Now that we have *glimpse*d our data and *select*ed our variables in the `starwars`, let's *arrange* them using by height. What function would we add to our code below to *arrange* our data so that the tallest person is listed at the top? (Hint: since `tidyverse` *arrange*s numbers from the smallest to the greatest number, you may need another function so the information is in *desc*ending order.)

```{r arrangeexercise, exercise = TRUE}
starwars %>% 
 select(name, height, homeworld, species)
```

<div id="selectexercise-hint">
**Hint:** I recommend checking out the `arrange()` function, which we talk about in [Chapter 3](https://utdata.github.io/rwdir/counts-import.html#arrange-the-data)
</div>

```{r arrangeexercise-solution}
starwars %>% 
  select(name, height, homeworld, species) |> 
  arrange(height %>% desc())
```

```{r arrangeexercise-check}
grade_this({
  if(identical(.result, .solution)){
    pass(sample(praise_list, 1))
  }
  fail()
})
```

So many tall characters! If your humanoid robot editor (J-327) wanted you to write a story about the tallest character in the dataset, who would that be?

```{r quiz-arrange, echo = FALSE}
quiz(
  question("Who is the tallest character?",
    answer("Yoda"),
    answer("Darth Vader"),
    answer("Chewbacca"),
    answer("Yarael Poof", correct = TRUE)
  )
)
```

Great job! Please inform your JedR Master that you have completed this exercise. 
