---
title: "JedR Knight Trail: Summarizing by Math"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
#library(gradethis)

starwars <- starwars
```

> THIS ONE COULD USE AN EDIT REVIEW

Hello again! In this exercise, you'll be practicing the skills you learned in Chapter 5 of [our textbook](https://utdata.github.io/rwdir/sums-analyze.html#item-quantities-totals-for-local-agencies).  Like our previous exercise, we'll be using the dataset `starwars` in the `tidyverse` packages.

## Summarizing

If you recall from our earlier exercise, `starwars` is a dataset with 87 Star Wars characters.

```{r glimpse}
glimpse(starwars)
```

In our earlier exercise, we counted the number of characters that were different species. In this exercise, we'll use `group_by()` again. But, instead of getting a count of the characters, let's get the **mean**, or average, height of each species.

```{r avgexercise, exercise = TRUE}
starwars %>%
  group_by(species)
```

<div id="gbsumexercise-hint">
**Hint:** Hmm..
</div>

```{r avgexercise-solution, exercise.reveal_solution = FALSE}
starwars %>%
  group_by(species) %>%
  summarize(height_avg = mean(height))
```

How would we arrange this information?
```{r arrangeexercise, exercise = TRUE}
starwars %>%
  group_by(species)
```

<div id="arrangeexercise-hint">
**Hint:** Don't forget that you want to sort this information in **desc**ending order, as we do in [Chapter 4](https://utdata.github.io/rwdir/count-analysis.html#arrange-the-results)
</div>

```{r arrangeexercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
starwars %>%
  group_by(species) %>%
  summarize(height_avg = mean(height)) %>%
  arrange(height_avg %>% desc())
```


```{r quiz-arrange, include = FALSE}
quiz(
  question("On average, what Star Wars species is the tallest?",
    answer("Wookiee"),
    answer("Droid"),
    answer("Quermian", correct = TRUE),
    answer("Ewok")
  )
)
```
