---
title: "JedR Knight Trail: Visualizing"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
#library(gradethis)

starwars <- starwars
```

> THIS ONE COULD USE AN EDIT REVIEW

Hello again! In this exercise, you'll be practicing the skills you learned in Chapter 7 of [our textbook](https://utdata.github.io/rwdir/ggplot-intro.html).  Like our previous exercise, we'll be using the dataset `starwars` in the `tidyverse` packages.

## Summarizing

If you recall from our earlier exercise, `starwars` is a dataset with 87 Star Wars characters.

```{r glimpse}
glimpse(starwars)
```

Like in our earlier exercise, let's group these characters by species and count the number of characters!

```{r speciesctexercise, exercise = TRUE}
starwars_count <- starwars %>%
```

<div id="speciesctexercise-hint">
**Hint:** If you need a refresher, check out [Chapter 4](https://utdata.github.io/rwdir/count-analysis.html#group-by).
</div>

```{r speciesctexercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
starwars_count <- starwars %>%
  group_by(species) %>% #groups the character by species
  summarize(count = n()) #count the number of characters of each species
```

## Plotting
Now that we have our 

```{r speciesplot, exercise = TRUE, exercise.setup = "speciesctexercise-solution"}
ggplot()
```

<div id="speciesplot-hint">
**Hint:** Don't forget to `reorder()` the x-axis! Check out [Chapter 7](https://utdata.github.io/rwdir/ggplot-intro.html#reorder-the-bars) for more.
</div>

```{r speciesplot-solution, exercise.reveal_solution = FALSE, include = FALSE}
ggplot(starwars_count, aes(x = reorder(species, count), y = count)) +
  geom_col() +
  coord_flip()
```
