---
title: "JedR Trial: Count"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(gradethis)

starwars <- starwars
gradethis_setup()

praise_list <- c("Great job you have done!", "Nice work. You have controlled your fear.", "Wow! The force is with you!", "Great job! Mind tricks don't work on you!", "The force is strong with this one.")

```


## JedR Trial No. 2: Summarize by counting

As a **JedR Padawan** of the *Galactic News Hub*, you have been assigned to help with a new story about the demographics of the Star Wars Universe. After pitching a few different topics, your droid editor (J-327D) has green lighted a pitch about species variation. For this article, you have been asked to provide a summary statistic about which species is the most common in the Star Wars universe.

For this task, you will **summarize** and **group** the characters by species. You will then **arrange** the data to identify the most common species of all the Star Wars characters.

### 1. Summarizing

If you recall from our earlier exercise, `starwars` is a dataset with 87 Star Wars characters.

```{r glimpse}
glimpse(starwars)
```

First, we'll need to *group* the data by species and then *summarize* by counting the *n*umber of rows of each species. (We will tackle the arranging part later.)

How do we do that? What functions would we use?

```{r gbsumexercise, exercise = TRUE}
starwars %>%
  
# we insist you name the new variable "characters"!
```

```{r gbsumexercise-solution}
starwars %>%
  group_by(species) %>%
  summarize(characters = n())
```

```{r gbsumexercise-check}
grade_this({
  if(identical(.result,.solution)){ 
    pass(sample(praise_list, 1))
  }
  fail()
})

```

<div id="gbsumexercise-hint">
**Hint:** Check out the `group_by()` and `summarize()` in [Chapter 4](https://utdata.github.io/rwdir/count-analysis.html#summarize)
</div>

### 2. Arrange the results

While the `group_by()` and `summarize()` functions are great for summarizing this information, a [Padawan youngling](https://www.starwars.com/news/what-is-a-padawan) has pointed out that the information is not ordered. 

🙄 Rolling your eyes, you tell them that you know this! We just need to use another function in order to **arrange** this data. To illustrate your point, let's **arrange** this data now.

As a tip, don't forget that we often want to sort the information by **desc**ending order, because journalists are typically interested in the *most* of something.

```{r arrangeexercise, exercise = TRUE, exercise.lines=5}
starwars %>%
  group_by(species) %>%
  summarize(characters = n())
```

<div id="arrangeexercise-hint">
**Hint:** Check out `arrange`. You can [learn more here](https://dplyr.tidyverse.org/reference/arrange.html).  
</div>

```{r arrangeexercise-solution}
starwars %>%
  group_by(species) %>%
  summarize(characters = n()) %>%
  arrange(characters %>% desc())
```

```{r arrangeexercise-check}
grade_this({
  if(identical(.result,.solution)){
    pass(sample(praise_list, 1))
  }
  fail()
})
```

Now that you've **grouped** the data, **summarized** the number of characters of each species, and **arranged** the data (in **desc**ending order), you can finally tell your editor that you've compiled this information!

```{r quiz-arrange, echo = FALSE}
question("What species is the most common in Star Wars?",
  answer("Human", correct = TRUE),
  answer("Droid"),
  answer("Aleena"),
  answer("Iktotchi"),
  random_answer_order = TRUE
)
```

Excellent! Please tell your JedR Master that you have completed this trial.
