---
title: "JedR: Managing dates"
tutorial:
  id: "jedr-manage-dates"
output:
  learnr::tutorial:
    progressive: true
    theme: lumen
    highlight: espresso
runtime: shiny_prerendered
---

##### Tutorial by: Christian Staal Bruun Overgaard, JedR trainer

## Introduction

Hello, young Padawan! In this tutorial we'll look at managing dates, often using `Lubridate` package, which can make life easier when dealing with dates in R. In doing so, we'll examine a dataset, scraped from [tvmaze.com](https://www.tvmaze.com/shows/117/star-wars-rebels/episodes), with information about the *Star Wars Rebels* series. I encourage you to follow along in your own R notebook! Good luck and may the force be with you!

```{r setup, warning=FALSE, message=FALSE, include=FALSE}
library(lubridate) # to work with dates and time
library(tidyverse) # to clean and wrangle the data
library(janitor) # to clean column names
library(learnr)
library(gradethis)
library(rsconnect)
library(praise)

### setup dataframes for the exercises
rebels_cl <- read_csv("https://raw.githubusercontent.com/utdata/jedr-academy/main/data/starwarsrebels.csv") |> 
  clean_names() |> 
  mutate(air_date_fixed = dmy(original_air_date)) |> 
  select(-original_air_date)

rebels_cl_newcols <- read_csv("https://raw.githubusercontent.com/utdata/jedr-academy/main/data/starwarsrebels.csv") |> 
  clean_names() |> 
  mutate(air_date_fixed = dmy(original_air_date)) |> 
  mutate(month = month(air_date_fixed, label = TRUE)) |> 
  mutate(year = year(air_date_fixed)) |> 
  select(-original_air_date)

parsing_practice_df <- rebels_cl |> 
  head(10) |>
  mutate(
    date2 = c("10-13-2014", "10-20-2014", "10-27-2014", "11-03-2014", "11-10-2014", "11-17-2014", "11-24-2014", "01-05-2015", "01-19-2015", "02-02-2015"),
    date3 = c("13-Oct-2014", "20-Oct-2014", "27-Oct-2014", "03-Nov-2014", "10-Nov-2014", "17-Nov-2014","24-Nov-2014", "05-Jan-2015", "19-Jan-2015", "02-Feb-2015"),
    date1 = as.character(air_date_fixed)) |> 
  select(episode_title, date1, date2, date3)
```

### Import data

```{r import, message=F}
rebels_raw <- read_csv("https://raw.githubusercontent.com/utdata/jedr-academy/main/data/starwarsrebels.csv") # import data
rebels_raw |> head() # see what the data looks like
```

This dataset has information about the *Star Wars Rebels* series. For each of the show's 69 episodes, we have its episode number, air date, title, and season.

Note that some of the column names have spaces in them, which can make life harder. To fix it, let's use the `clean_names()` function from the `Janitor` package:

```{r clean_names}
rebels <- rebels_raw |> clean_names() # clean the names
rebels |> glimpse() # glimpse the "rebels" object
```

Now, the column names are easier to work with.

## Parsing dates: How to reconize dates in R

Notice that when we used the `glimpse()` above, we also got info about the type of each column. The episode number and season are read as *dbl* ("doubles, which is a number"), whereas the episode title and air date are read as *chr* ("characters").

To work effectively with dates, we need R to recognize when we're reading with dates. Do fix this, we can use `lubridate`'s [parse date functions](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf). In this particular dataset, the dates are written in the *dmy* (date-month-year) format. (If you look in the `glimpse()` output above, you'll see that the first observation, for example, was: "13-Oct-14"). Let's use `lubridate`'s `dmy()` function to create a new version of the dataset, called *rebels_fix*, which has a new column called *air_date_fixed*:

```{r glimpse}
rebels |> 
  mutate(air_date_fixed = dmy(original_air_date)) |> 
  glimpse()
```

Notice that, in the chunk above, we're just looking at the output to see that this worked as intented. See that the `air_date_fixed` column is being read as a `date` whereas the `original_air_date` column was read as a `character`? Also, we're not saving it into a new object using the `<-` operator. Let's use the `<-` operator operator to save our work in a new object, called `rebels_cl`, without the `original_air_date` column, which we're not going to need:

```{r arrange}
rebels_cl <- rebels |> 
  mutate(air_date_fixed = dmy(original_air_date)) |> 
  select(-original_air_date)

rebels_cl |> glimpse()
```

Two key things to notice here:

* `air_date_fixed` has the format `date`.
* In `air_date_fixed`, the year is written first, followed by the month and then the day. That's just what the date format looks like in R.

In this example, `lubridate`'s `dmy()` function came in handy. Here are some other functions we might've used if the dates had been formatted differently:

```{r mdy_ymd}
### when using the mdy function, R knows that the first value is month, followed by day, followed by year
mdy("May 15, 1988")

### here, the ymd function makes it clear that the first value is year, followed by month, followed by day
ymd("1988-May-15")
```

You can find more examples and tips [in this cheatsheet](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf).

#### Padawan practice: Parsing dates

```{r make_parsing_practice_df, include=FALSE}
parsing_practice_df <- rebels_cl |> 
  head(10) |>
  mutate(
    date2 = c("10-13-2014", "10-20-2014", "10-27-2014", "11-03-2014", "11-10-2014", "11-17-2014", "11-24-2014", "01-05-2015", "01-19-2015", "02-02-2015"),
    date3 = c("13-Oct-2014", "20-Oct-2014", "27-Oct-2014", "03-Nov-2014", "10-Nov-2014", "11-17-2014","24-Nov-2014", "05-Jan-2015", "19-Jan-2015", "02-Feb-2015"),
    date1 = as.character(air_date_fixed)) |> 
  select(episode_title, date1, date2, date3)
```

Here's a small portion of the data, called `parsing_practice_df`, with three columns that need to be parsed:

```{r pars_prac}
parsing_practice_df
```

Each column is currently written in the *character* format and should be converted to the *date* format. Can you figure out how to do this?

##### Practice: Parse `date1`

*Hint: Replace the three lines, `___` with the appropriate function.*

```{r parsing_practice1a-exercise, exercise = TRUE}
parsing_practice_df |> 
  mutate(date1_cleaned = ___(date1))
```

```{r parsing_practice1a-exercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
parsing_practice_df |> 
  mutate(date1_cleaned = ymd(date1))
```

```{r parsing_practice1a-exercise-check}
grade_this_code(correct = paste("Correct, young Padawan! We're using the ymd() function to indicate that the year goes first, followed by month, followed by day. This is our way of telling R how to parse this date."))
```

Here, as in the two following exercises, we're not saving the output but only displaying it. If we wanted to save it, we'd use the `<-` operator. Okay, see if you can solve the two following exercises, too!

##### Practice: Parse `date2`

```{r parsing_practice2a-exercise, exercise = TRUE}
parsing_practice_df |> 
  mutate(date2_cleaned = ___(date2))
```

```{r parsing_practice2a-exercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
parsing_practice_df |> 
  mutate(date2_cleaned = mdy(date2))
```

```{r parsing_practice2a-exercise-check}
grade_this_code(correct = paste("Bravo, Padawan! Looks like the new column is in the date format!"))
```

##### Practice: Parse `date3`

```{r parsing_practice3a-exercise, exercise = TRUE}
parsing_practice_df |> 
  mutate(date3_cleaned = ___(date3))
```

```{r parsing_practice3a-exercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
parsing_practice_df |> 
  mutate(date3_cleaned = dmy(date3))
```

```{r parsing_practice3a-exercise-check}
grade_this_code(correct = paste("Correct! The force is strong in you! Note that this function works regardless of whether the months in the existing column are written out using characters or numbers"))
```

### Another way to parse dates

Base R also has a function to parse dates, called `as.Date()`, which you can read more about [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/as.Date).

## Working with dates

Let's now make use of our newly created `air_date_fixed` column to understand the data.

### Which episodes were oldest or newest?

We can use the `arrange()` function to sort data frames based on dates.

#### The oldest episodes

```{r arrange2}
rebels_cl |> 
  arrange(air_date_fixed)
```

#### The newest episodes

```{r newest}
rebels_cl |> 
  arrange(desc(air_date_fixed)) # add "desc" to sort in descending order
```

### Only view episodes that aired before or after a given date

We can use the `filter()` function to only view episodes that aired before or after a given date:

#### Episodes aired in 2018 (the last year available)

```{r datefix}
rebels_cl |> 
  filter(air_date_fixed >= "2018-01-01") #Notice that the date is written in quotes!
```

#### Episodes that aired in February 2016

```{r filtering}
rebels_cl |> 
  filter(air_date_fixed >= "2016-02-01" & air_date_fixed < "2016-03-01")
```

#### Padawan practice: using filter() with dates

Identify the episodes that aired in October 2017. 

Hint: Use the `filter()` function to specify that you only want to include episodes that aired on October 1, 2017 or later—and also aired before November 1 the same year.

```{r filter-exercise, exercise = TRUE}
rebels_cl |> 
  filter() # fill out the filter function!
```

```{r filter-exercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
rebels_cl |> 
  filter(air_date_fixed >= "2017-10-01" & air_date_fixed < "2017-11-01")
```

```{r filter-exercise-check}
#that_correct <- c('Good job, Padawan!')
#grade_this_code(correct = paste(that_correct, praise() ) )
grade_this_code(correct = paste("That's right! The force is strong with you!"))
```

## Using specific parts of a date

### Which month had the most episodes (throughout the entire period)?

To answer this question, we can use `Lubridate`'s `month()` function, which allows us to focus only on months (while ignoring years). It works like this:

```{r month}
rebels_cl_newcols <- rebels_cl |> 
  mutate(month = month(air_date_fixed))

rebels_cl_newcols |> glimpse()
```

Here, we're creating a new dataframe called `rebels_cl_newcols`. We'll use this to add more columns later on.

It's worth keeping in mind that "month" is now the name of a column (which we just created) *and* the name of a function from the `lubridate` package. That can be a bit confusing and it's generally best to avoid creating object with the same names as functions. 

Also, see how the newly created `month` column only has information about which month the episodes aired in, in this case "10" (October) for the first three observations and "11" (November) for the next four observations. If we wanted the names of the months, we can use the `label = TRUE` argument within the month function. Like this:

```{r month_label}
rebels_cl_newcols <- rebels_cl_newcols |> 
  mutate(month = month(air_date_fixed, label = TRUE))

rebels_cl_newcols |> glimpse()
```

Notice that the type for *month* is *"ord"*? This means that the column is written as an "ordered factor", which simply means (in this case) that although it looks like text, it would be ordered by time, not alphabetically. In other words, we can sort by the month column, if we want to do so:

```{r sort_month}
rebels_cl_newcols |> 
  select(episode_title, month) |> 
  arrange(month)
```
See how it's arranged by month?


Going back to our question about which month had the most episodes, let's take a look:

```{r counting}
rebels_cl_newcols |> 
  count(month, name = "number_of_episodes") |> 
  arrange(desc(number_of_episodes) )
```

It looks like there were lot's of new episodes in October, November, February, and March. 

As a quick aside, when we do the month labels, we can also get R to write out the full month rather than the abbreviation. Let's look at an example:

```{r maythefourth}
month("2022-05-04", label = TRUE, abbr = FALSE)
```

The `abbr` argument defaults to TRUE, so if we don't specific `abbr = FALSE`, we'd get the abbreviated version as we saw above.

#### Padawan practice: using the `year()` function

```{r making_it_run_in_rstudio, include=FALSE}
rebels_cl_newcols <- rebels_cl |> 
  mutate(year = year(air_date_fixed)
           )
```

To get information about specific years, we can use the `year()` function, which is quite similar. Try to create a variable called `year` see we can check how many episodes aired in each year.

Hint: you'll need to use the `air_date_fixed` column.

```{r month-function-exercise, exercise = TRUE}
### USE THE year() FUNCTION HERE TO CREATE A NEW VARIABLE CALLED "year"
rebels_cl_newcols <- rebels_cl_newcols |> 
  mutate( )

### finding number of episodes per year (you don't need to change the code below)
rebels_cl_newcols |> 
  group_by(year) |> 
  count(name = "number_of_episodes") |> 
  arrange(desc(number_of_episodes) )
```

```{r month-function-exercise-solution, exercise.reveal_solution = FALSE, include = FALSE}
rebels_cl_newcols <- rebels_cl_newcols |> 
  mutate(year = year(air_date_fixed)
           )

### finding number of episodes per year (you don't need to change the code below)
rebels_cl_newcols |> 
  group_by(year) |> 
  count(name = "number_of_episodes") |> 
  arrange(desc(number_of_episodes) )
```

```{r month-function-exercise-check}
that_correct <- c('Correct, Padawan!')
#master_soon <- c('A JedR master soon you will be!')
grade_this_code(correct = paste(that_correct, praise()) )
```


### Floor_date()

An alternative would be to use [`Lubridate`'s `floor_date()`-function](https://lubridate.tidyverse.org/reference/round_date.html), which let's us round date-time objects *down* to their nearest year, month, week, day, etc. It works like this:

```{r floor_date}
rebels_cl_newcols |> 
  mutate( #use the mutate function to create a new column
    month_floor = #set the new column name to 
      floor_date(air_date_fixed, #use the round_date function on our existing date varible
                       "month")) |> #we want to round to nearest month
  select(episode_title, month_floor) |> #we only want to view these columns now
  head(10) #print just the ten first rows
```

Notice that the new column, *month_floor*, only include dates that are the first day in any given month? 

Rounding the dates to their nearest year/month/week/day/etc can be helpful later on, for example when visualizing the data. Instead of round to the nearest values, we can use `ceiling_date()` to round up to the nearest value or `round_date()` to round to the closest value. In this tutorial we'll mostly focus on `floor_date()` but as a future JedR Master, you should know about the other options, too! 

Okay, let's create a dataframe with four new columns, rounded by `floor_date()`. We'll use this later on.

```{r rebels_rounded_cols}
rebels_rounded_cols <- rebels_cl_newcols |> 
  mutate(year_fl = floor_date(air_date_fixed, "year")) |>
  mutate(month_fl = floor_date(air_date_fixed, "month")) |>
  mutate(week_fl = floor_date(air_date_fixed, "week")) |>
  mutate(day_fl = floor_date(air_date_fixed, "day"))

rebels_rounded_cols |> select(episode_title, year_fl, month_fl, week_fl, day_fl) |> glimpse()
```

### The yday() function

What if we wanted to find the earliest date within a year that an episode aired? For that purpose, we could use the `yday()` (year day) function, which simply tells us how many days into a year a given date is. My son and I share the same birthday (February 5). That's the 36th day of the year:

```{r yday}
yday("2022-02-05")
```

To find the earliest episode airdate for each year, we could do this:

```{r dy_chunk}
rebels_rounded_cols |>
  mutate(
    dy = yday(air_date_fixed),
    year = year(air_date_fixed)) |> 
  group_by(year) |> 
  slice_min(air_date_fixed) |> 
  ungroup() |>
  select(episode_title, season, air_date_fixed, year, dy)
```

This shows the first-aired episode for each of the years from 2014–2018. Notice the few doppelgängers? It looks like seasons 3 and 4 started each started with a double episode.

There are few things going on in the chunk above. Here's a quick recap of what each line does:

* We start with our object, `rebels_rounded_cols`, and then ...
* We use `mutate()` to create a new column called `dy` ...
* To get the `dy` column to specify each dates number within a year, we use `yday()`, and then ...
* We `group_by()` because we want the earliest episode within each year, and then ...
* We use `slice_min()` to get just the earliest episode within each year. (Otherwise, our output would all the episodes.) And then ...
* We use `ungroup()` to remove the grouped `dy` column. Not essential here but might be handy if you need to perform additional operations. And then ...
* `select()` let's us show just a few columns to make the output more readable at a glance.

## Plotting with dates

When working with dates it can be super useful to plot your date. This can show how things are changing over time. Let's look at a quick example. Remember above where we used `floor_date()` to round all dates down to their nearest month? We can use that column to plot how many episodes that aired per month throughout the entire period.

First, let's get the data into a more handy format.

```{r ggplot-prep-gs}
rebels_gs <- rebels_rounded_cols |> 
  group_by(month_fl) |> 
  summarise(no_episodes = n())

rebels_gs |> head() #let's see what this new dataframe looks like
```

A quick rundown of what's happening above:

* We're grouping by month_fl. This is because we're interested in the number of episodes per month. We could also have used year_fl, week_fl, or day_fl here.
* Then we're using the summarise function to simply count the number of episodes per month.
* We're storing this in a new dataframe, `rebels_gs`, which we'll use below.

With that taken care of, here's an example of how we could use `ggplot` to visualize it:

```{r actual-plot}
ggplot(rebels_gs, aes(x = month_fl, y = no_episodes)) + 
  geom_bar(stat='identity') +
  labs(
        subtitle = "Number of Star Wars Rebels episodes per month",
        caption = "JedR Masters Lubridate Tutorial") + 
  ggthemes::theme_fivethirtyeight()
```

Here's a quick explanation of what's happening in the chunk above:

* In the first line, we're specifying the dataset ("rebels_gs", x-variable ("month_fl"), and y-variable ("no_episodes")
* In the second line, we're using `geom_bar()` to tell R to make a bar chart. We use the "stat = 'identity'" arugment to clarify that we'll be providing the values for the bars
* In the third portion, we're using `labs()` to create the title and caption. We're using "subtitle" rather than "title" b/c it works better with the theme we're using
* In the fourth and final portion, we're employing `theme_fivethirtyeight()` from the `ggthemes` package.

## The end

That's it for now, young Padawan! Good luck on your next quest, stay safe, and may the force be with you! 💪
